Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using Multiple Automatically Generated Templates

### 2013-09-03 paper.Rnw

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has 
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
Multiple Automatically Generated Templates}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1,8]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4,5]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,6,8]{Aristotle N. Voineskos}
\author[1,6,7]{M. Mallar Chakravarty}
\author[ ]{and the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.ucla.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}
\affil[8]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

The hippocampus is a brain structure situated in the medial temporal
lobe, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. The hippocampus is of interest to
neuroscientists because it is implicated in many forms of brain disfunction,
including Alzheimer's disease \citep{Sabuncu2011} and schizophrenia
\citep{Narr2004,Karnik-Henry2012}. In neuroimaging studies, structural magnetic
resonance images (MRI) are often used for the volumetric assessment of the
hippocampus.  As such, accurate segmentation of the hippocampus and its
subregions in MRI is a necessary first step to better understand the
inter-individual variability of subject neuroanatomy.

@@

In this paper we undertake a thorough validation of the \mb algorithm for
segmentation of the hippocampus. First, we address the feasibility and accuracy
of whole hippocampus segmentation with a limited number of input atlases by
performing a multi-fold experiment using a subset of the Alzheimer's Disease
Neuroimaging Initiative (ADNI) 1 dataset. Based on the parameters we find in our
experiment, we then validate our algorithm using all of the data available in
the ADNI Complete 1Yr sample and compare our segmentations to the other
segmentations that are available through the ADNI informatics portal. To ensure
that we have not over-fit our parameters to the aging or neurodegenerative
brain, we also apply our segmentations to a dataset of individuals suffering
from first episode psychosis. Finally, we perform a leave-one-out validation
experiment to determine if the subfields can be accurately identified using our
multi-atlas framework.

@@

\subsection{The \mb Algorithm}
In this paper, we will use the term {\em label} to mean any segmentation (manual or 
derived) of an MR image. {\em Label propagation} is the process by which two
images are registered and the resulting transformation is applied to the labels
from one image to bring them into alignment with the other image. We will use
the term {\em atlas} to mean a manually segmented image, and the term {\em
template} to mean an automatically segmented image (i.e.  via label
propagation). The terms {\em atlas library} and {\em template library} describe
any set of such images. Additionally, we will use the term {\em target} to refer
to an unlabelled image that is undergoing segmentation.

The simplest form of multi-atlas segmentation, (which we will call {\em basic 
multi-atlas segmentation}), involves three steps. First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single label by label fusion
\citep{Heckemann2006, Heckemann2011}. The basic multi-atlas segmentation method
is described in detail in other publications
\citep{Collins2010,Heckemann2011,Aljabar2009}. When only a single atlas is used,
basic multi-atlas segmentation degenerates into model-based segmentation: labels
are propogated from the atlas to a target, and no label fusion is needed.

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. The last two experiments assess the
validity of the \mb algorithm when applied to different diseases: first episode
schizophrenia (Experiment 3), and Alzheimer's disease (Experiment 4).
Additionally, in Experiment 4, we compare \mb segmentations with those of
well-known automated methods and assess segmentation bias.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Semi-automated segmentations of the
left and right whole hippocampi are made available with a subset of ADNI images
\citep{Hsu2002}. These labels have been generated using the SNT tool from
Medtronic Surgical Navigation Technologies, Louisville, CO (see Supplementary
Materials for detailed discussion of the segmentation process used). Throughout
this paper, we will use these segmentations as the \gs for comparison.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<fig:ADNI1-xval-k-mean, cache=TRUE, dependson='ADNI1-xval-prep', fig.width=7, fig.height=7>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(all_data_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=as.factor(atlases)), width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Number of Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  opts(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \caption{Mean Dice's Similarity Coefficient of \mb segmentations relative to
  SNT \gs segmentations for 69 ADNI1 subjects vs. atlas and template library
  size, registration algorithm, and label fusion method. Error bars indictate
  standard error.
  \label{fig:ADNI1-xval-k-mean}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Figure \ref{fig:ADNI-segmentations} shows a qualitative comparison of \mb and 
SNT \gs hippocampal segmentations for 10 randomly selected subjects in 
each disease category, and illustrates some of the common errors found during 
visual inspection. Mostly frequently, we find \mb improperly includes the vestigial
hippocampal sulcus and, although not anatomically incorrect, \mb under-estimates 
the hippocampal body in comparison to the SNT \gs segmentation.

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=6in]{figure/ADNI1_SNT_MB_montage/montage.pdf}
  \end{centering}
  \caption{SNT \gs and \mb segmentations for 30 ADNI1 subjects (10
  subjects randomly selected from each disease category in the subject pool used
  in Experiment 1). Sagittal slices are shown for each unlabelled T1-weighted
  anatomical image. SNT \gs labels appear in green, and \mb labels appear in
  blue. Noted are examples of common segmentation idiosyncrasies: 
  {\em (a)} over-estimation of hippocampal head and  
  {\em (b)} translated SNT \gs segmentation; 
  {\em (c)} under-estimation of hippocampal body and
  {\em (d)} improper inclusion of the vestigial hippocampal sulcus by \mb.}
  \label{fig:ADNI-segmentations}
\end{figure}

@@

To this end, more work is required to determine the source of the slight
decrease in segmentation performance when the number of templates are set to an
even number.  Our initial concern was that this dip in performance was a
by-product of the \mb algorithm itself.  However, this this pattern is also
found in the results of the multi-atlas segmentations we used in our
experiments.  We believe that our majority voting methodology is biased towards
labels with the lowest numeric values when breaking ties (by way of the
implementation of the \verb+mode+ function used to determine majority), thus
causing the slight bias observed when using an even number of templates. This is
another area where the voting scheme could be used to improve performance.
However, it is worth noting that this limitation was previously identified by
\citet{Heckemann2006a} and, subsequently, other groups have not even considered
the potential pitfalls of an even number of candidate labels (eg.
\citet{Leung2010}).

Another concern is the moderate-to-small improvement observed in \mb in
comparison to multi-atlas segmentation when using the same number of atlases.
The actual benefit in using \mb is consistency of the labelling regardless of
atlas or template choice. This is an important consideration that few have
touched on previously. The 10-fold Monte Carlo leave-one-out cross-validation
that we present in Experiment 1 is amongst one of the most stringent performed
in the multi-atlas/segmentation literature. To the best of our knowledge, other
groups using ADNI data for validation do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}), aside from
\citep{Wang2011} who perform a similarily rigorous validation as us. The
thoroughness of our validation suggests that our results are reflective of a
true average over the choice of parameter settings and are independent of atlas
or template choice (provided the input atlases are consistently segmented).

On that note, it is worth observing that we found the SNT segmentations to be
poor in general. As previously noted in Figure \ref{fig:ADNI-segmentations}, an
expert manual rater (JW) identified many inconsistencies: occurances of over-
and under-estimation, as well as misalignments of the entire segmentation
volume.  Therefore, the use of the SNT segmentations as gold standards in
Experiments 1 and 4 presents the possibility that the DSC scores and
correlations we obtained are lower than what would be found with a more accurate
and consistent gold standard.

@@

\section*{R Session Information}
<<results = "asis">>=
toLatex(sessionInfo(), locale=FALSE)
@
\end{document}

>>>>

% %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% TODO: 
%  - check tense: we are is have has  use/used find/found
% Sweave/Knitr init
<<setup, include=FALSE, cache=FALSE>>=
library(knitr)
opts_chunk$set(echo = FALSE, 
               message = FALSE, 
               warning = FALSE, 
               fig.align = 'center', 
               tidy = FALSE, 
               comment = NA, 
               cache = TRUE,
               fig.width = 6, 
               fig.height = 6)
library(ggplot2)
library(Hmisc)
library(reshape2)
library(ADNIMERGE)
library(plyr)
options(digits=3)
theme_set(theme_bw(base_size = 12))

@@

% title and authors
\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
Multiple Automatically Generated Templates}
\author[1]{Jon Pipitone}
\author[1]{Matt T. Park}
\author[1]{Julie Winterburn}
\author[1,8]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4,5]{Jens C. Pruessner}
\author[4]{Martin Lepage}
\author[1,6,8]{Aristotle N. Voineskos}
\author[1,6,7]{M. Mallar Chakravarty}
\author[ ]{the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.ucla.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health, Hospital for Sick Children, Toronto,
ON,Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomedical Engineering, University of Toronto, Toronto,
ON, Canada}
\affil[8]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

The hippocampus is a brain structure situated in the medial temporal
lobe, and has long been associated with learning and memory
\citep{DenHeijer2012,Scoville2000}. The hippocampus is of interest to
neuroscientists because it is implicated in many forms of brain dysfunction,
including Alzheimer's disease \citep{Sabuncu2011} and schizophrenia
\citep{Narr2004,Karnik-Henry2012}. In neuroimaging studies, structural magnetic
resonance images (MRI) are often used for the volumetric assessment of the
hippocampus.  As such, accurate segmentation of the hippocampus and its
subregions in MRI is a necessary first step to better understand the
inter-individual variability of subject neuroanatomy.

@@

In this paper we undertook a thorough validation of the \mb algorithm for
segmentation of the hippocampus. First, we addressed the feasibility and
accuracy of whole hippocampus segmentation with a limited number of input
atlases by performing a multi-fold experiment using a subset of the Alzheimer's
Disease Neuroimaging Initiative (ADNI) dataset over a range of parameters.  We
then performed a leave-one-out validation to determine if hippocampal subfields
can be accurately identified using our multi-atlas framework.  To ensure that we
have not over-fit our parameters to the aging or neurodegenerative brain, we
also applied our segmentations to a dataset of individuals suffering from first
episode psychosis.  Finally, validated our algorithm using all of the data
available in the \adnidataset sample and compare our segmentations to the other
segmentations available through the ADNI informatics portal.

@@

\subsection{The \mb Algorithm}
In this paper, we use the term {\em label} to mean any segmentation (manual or
derived) of an MR image. {\em Label propagation} is the process by which two
images are registered and the resulting transformation is applied to the labels
from one image to bring them into alignment with the other image. We use the
term {\em atlas} to mean a manually segmented image, and the term {\em template}
to mean an automatically segmented image (i.e.  via label propagation). The
terms {\em atlas library} and {\em template library} describe any set of such
images. Additionally, we use the term {\em target} to refer to an unlabelled
image that is undergoing segmentation.

The simplest form of multi-atlas segmentation, which we call {\em basic 
multi-atlas segmentation}, involves three steps. First, each labelled 
input image (i.e. atlas or template) is registered to an unlabelled target image.
Second, the labels from each image are propagated to the target image space. 
Third, the labels are combined into a single label by label fusion
\citep{Heckemann2006, Heckemann2011}. The basic multi-atlas segmentation method
is described in detail in other publications
\citep{Collins2010,Heckemann2011,Aljabar2009}. When only a single atlas is used,
basic multi-atlas segmentation degenerates into model-based segmentation: labels
are propagated from the atlas to a target, and no label fusion is needed.

@@

The following section describes for experiments conducted to assess the
segmentation quality of the \mb algorithm. The first two experiments assess the
validity of \mb using a cross-validation design. Experiment 1 investigates the
accuracy of \mb whole hippocampus segmentation over a wide range of parameter
settings. The results of this experiment enable us to choose the parameter
settings offering the best performance for use in subsequent experiments.
Experiment 2 assesses hippocampal subfield segmentation quality in a
leave-one-out cross-validation design. The last two experiments assess the
validity of the \mb algorithm when applied to different diseases: first episode
schizophrenia (Experiment 3), and Alzheimer's disease (Experiment 4).
Additionally, in Experiment 4, we compared \mb segmentations with those of
well-known automated methods and assessed segmentation bias.

@@

69 1.5T images were arbitrarily selected from the baseline scans in the {\em
\adnidataset} standardized dataset. 23 subjects were chosen from each disease
category: cognitively normal (CN), mild cognitive impairment (MCI) and
Alzheimer's disease (AD). Demographics for this subset are shown in Table
\ref{tab:ADNI1-xvalidation-demographics}.  Semi-automated segmentations of the
left and right whole hippocampi are made available with a subset of ADNI images
\citep{Hsu2002}. These labels have been generated using the SNT tool from
Medtronic Surgical Navigation Technologies, Louisville, CO (see Supplementary
Materials for detailed discussion of the segmentation process used). Throughout
this paper, we use these segmentations as the \gs for comparison.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
\begin{figure}
<<fig:ADNI1-xval-k-mean, cache=TRUE, dependson='ADNI1-xval-prep', fig.width=7, fig.height=7>>=
stderr <- function(x) sqrt(var(x)/length(x))
ggplot(subset(all_data_mean, (templates.mb*atlases) %% 2 == 1),
       aes(x=templates.mb,y=k.mb, colour=as.factor(atlases))) + #as.factor((atlases*templates.mb) %% 2 ))) + 
  facet_grid(reg_method~method.mb) + 
  stat_summary(fun.y=mean,geom='line',
               aes(y=k.mb, weight=1, group=as.factor(atlases))) + 
  stat_summary(fun.y=mean,
               fun.ymin=function (x) mean(x) - stderr(x),
               fun.ymax=function (x) mean(x) + stderr(x), 
               geom='errorbar',
               aes(y=k.mb, weight=1, group=as.factor(atlases)), width=0.5) + 
  #stat_smooth(method='lm', formula=y~log(x)) + 
  scale_colour_hue(name="Number of Atlases") +
  xlab( "Number of Templates" ) + 
  ylab( "Mean DSC" ) + 
  opts(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text = element_text(size = 8))
@
  \caption{Mean Dice's Similarity Coefficient of \mb segmentations relative to
  SNT \gs segmentations for 69 ADNI subjects vs. atlas and template library
  size, registration algorithm, and label fusion method. Error bars indicate
  standard error.
  \label{fig:ADNI1-xval-k-mean}}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Figure \ref{fig:ADNI-segmentations} shows a qualitative comparison of \mb and 
SNT \gs hippocampal segmentations for 10 randomly selected subjects in 
each disease category, and illustrates some of the common errors found during 
visual inspection. Mostly frequently, we found \mb improperly includes the vestigial
hippocampal sulcus and, although not anatomically incorrect, \mb under-estimates 
the hippocampal body in comparison to the SNT \gs segmentation.

@@

\begin{figure}
  \begin{centering}
    \includegraphics[width=6in]{figure/ADNI1_SNT_MB_montage/montage.pdf}
  \end{centering}
  \caption{SNT \gs and \mb segmentations for 30 ADNI subjects (10
  subjects randomly selected from each disease category in the subject pool used
  in Experiment 1). Sagittal slices are shown for each unlabelled T1-weighted
  anatomical image. SNT \gs labels appear in green, and \mb labels appear in
  blue. Noted are examples of common segmentation idiosyncrasies: 
  {\em (a)} over-estimation of hippocampal head and  
  {\em (b)} translated SNT \gs segmentation; 
  {\em (c)} under-estimation of hippocampal body and
  {\em (d)} improper inclusion of the vestigial hippocampal sulcus by \mb.}
  \label{fig:ADNI-segmentations}
\end{figure}

@@

To this end, more work is required to determine the source of the slight
decrease in segmentation performance when the number of templates are set to an
even number.  Our initial concern was that this dip in performance was a
by-product of the \mb algorithm itself.  However, this this pattern is also
found in the results of the multi-atlas segmentations we used in our
experiments.  We believe that our majority voting methodology is biased towards
labels with the lowest numeric values when breaking ties (by way of the
implementation of the \verb+mode+ function used to determine majority), thus
causing the slight bias observed when using an even number of templates. This is
another area where the voting scheme could be used to improve performance.
However, it is worth noting that this limitation was previously identified by
\citet{Heckemann2006a} and, subsequently, other groups have not even considered
the potential pitfalls of an even number of candidate labels (e.g.
\citet{Leung2010}).

Another concern is the moderate-to-small improvement observed in \mb in
comparison to multi-atlas segmentation when using the same number of atlases.
The actual benefit in using \mb is consistency of the labelling regardless of
atlas or template choice. This is an important consideration that few have
touched on previously. The 10-fold Monte Carlo leave-one-out cross-validation
that we present in Experiment 1 is amongst one of the most stringent performed
in the multi-atlas/segmentation literature. To the best of our knowledge, other
groups using ADNI data for validation do at most a single round of
leave-one-out-validation (Table \ref{tab:other-methods}), aside from
\citep{Wang2011} who perform a similarly rigorous validation as us. The
thoroughness of our validation suggests that our results are reflective of a
true average over the choice of parameter settings and are independent of atlas
or template choice (provided the input atlases are consistently segmented).

On that note, it is worth observing that we found the SNT segmentations to be
poor in general. As previously noted in Figure \ref{fig:ADNI-segmentations}, an
expert manual rater (JW) identified many inconsistencies: occurrences of over-
and under-estimation, as well as misalignments of the entire segmentation
volume.  Therefore, the use of the SNT segmentations as gold standards in
Experiments 1 and 4 presents the possibility that the DSC scores and
correlations we obtained are lower than what would be found with a more accurate
and consistent gold standard.

@@

% \section*{R Session Information}
% <<results = "asis">>=
% toLatex(sessionInfo(), locale=FALSE)
% @
\end{document}
